Joint maximum-likelihood of phylogenies and ancestral states is not consistent

### 2018-07-25 appendix.tex
consistency graf

>>>>

\subsection*{Convergence of branch parameters}

%das: @EM I think this is actually the same Wald reference that Felsenstein tried to use.
%It works here for a fixed topology, and didn't work for him or your reference because I think the upper-semicontinuous condition doesn't hold for trees in general.
%Also I don't think the compactness condition holds---tree space isn't compact is it?
%However we now have a potential snag in that we are doing what Yang (1994) in his intro says is bad, i.e., doing ML for fixed, separate topologies and comparing them.
%We still have the behavior in the plots and the proofs, since they all assumed a fixed topology, and I think everything's fine with the arguments about Fels vs InvFels, but maybe you can let me know what you think?
For a fixed $\tau$, we show that $\hat{t}_\nCols \rightarrow \hat{t}$ for
\[
\hat{t}_\nCols = \arg\max_{t\in\mathcal{T}} \ \frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment)
\]
and
\[
\hat{t} = \arg\max_{t\in\mathcal{T}} \ \ell_{\tau^*,t^*}(\tau, t).
\]
Using the notation in \citet{van1998asymptotic}, we let
\[
m_t(\alignmentColumn) = \sum_{j=1}^\nSiteSplits 1\{\patternToSplit(\alignmentColumn)=\siteSplit_j\} \cdot [\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]
\]
so that
\[
\frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment) = \frac{1}{n} \sum_{i=1}^\nCols m_t(\alignmentColumn_i)
\]
and
\[
\ell_{\tau^*,t^*}(\tau, t) = E[m_t].
\]
To show $\hat{t}_\nCols \rightarrow \hat{t}$, we use Wald's consistency proof \citep[p. 48, Theorem 5.14 of ][]{van1998asymptotic}, which requires four conditions.
The first is that $\mathcal{T}$ is compact, which is obviously true.
The second is that
\[
E\left[\sup_{t\in\mathcal{T}} m_t\right] < \infty,
\]
and, since $m_t(\alignmentColumn)$ is nonpositive for all $t$ and $\alignmentColumn$, this property holds.
The remaining conditions are on the maps
\[
\alignmentColumn \mapsto \sup_t m_t(\alignmentColumn)
\]
and
\[
t \mapsto m_t(\alignmentColumn).
\]
We need the first map to be measurable, which is evident since the domain $\alphabet^\nSiteRows$ of the mapping is a finite set, and so all subsets of the domain are also finite and thus measurable.
Finally, we must have the the second mapping be upper-semicontinuous for almost all $\alignmentColumn$.
For a fixed ancestral state split $t \mapsto m_t(\alignmentColumn)$ is continuous for all $\alignmentColumn$.
Discontinuities occur when, if we move about in $\mathcal{T}$, a different ancestral state split becomes more likely, though when we are at a point of discontinuity we maximize over ancestral state splits, ensuring the upper-semicontinuous property of this mapping.
This shows $\hat{t}_\nCols \rightarrow \hat{t}$, allowing our consistency results to be proved using $\ell_{\tau^*,t^*}(\tau, t)$.

### 2018-07-25 joint_inf.tex
consistency graf

Assume $\fullAlignment$ is generated from the InvFels topology $\tau^*$ and with true generating branch fidelities $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ (Fig.~\ref{fig:farris-fels-top}a).
Use $\ell_{\tau^*,t^*}(\tau, t)$ to denote the expected per-site log-likelihood, which can be thought of as the infinite-length sequence case
\begin{equation}
\label{eq:exp-log-lik}
\frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment) \rightarrow \ell_{\tau^*,t^*}(\tau, t).
\end{equation}
%EM Could we give an extra sentence justifying that the limit of the inferential process will be the same as the inferences applied to the limiting data? Compare https://paperpile.com/shared/HohiVH in which they are rather different.
%das: wow, good catch! This is not at all obvious to me it turns out :/ I guess I'll have to think harder on it
We give $\ell$ explicitly as \eqref{eq:site_pattern_profile_likelihood_mean} in the Appendix.

@@

%EM Suggest "Our main theorem states that ... internal branch length zero ..."
Our main theorem states that given data generated on $\tau_1$ there exist true nonzero branch lengths such that the estimator $\hat{t}$ maximizing \eqref{eq:exp-log-lik} has an internal branch of length zero.
\begin{restatable}{theorem}{topoInconsist}
Let $\tau^*=\tau_1$, $t^*=\{x^*, y^*, x^*, y^*, y^*\}$, and $t=\{x_1, y_1, x_2, y_2, w\}$ with $x_1, y_1, x_2, y_2, w > 0$.
There exists a set of $0 < x^*, y^* < 1$ such that the solution $\hat{t} := \{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\}$ given by
\[
\hat{t} = \arg\max_{t} \ \ell_{\tau^*,t^*}(\tau_1, t)
\]
has the property $\hat{w}\equiv 1$.
\end{restatable}

>>>>

Assume $\fullAlignment$ is generated from the InvFels topology $\tau^*$ and with true generating branch fidelities $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ (Fig.~\ref{fig:farris-fels-top}a).
Use $\ell_{\tau^*,t^*}(\tau, t)$ to denote the expected per-site log-likelihood, which can be thought of as the infinite-length sequence case
\begin{equation}
\label{eq:exp-log-lik}
\frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment) \rightarrow \ell_{\tau^*,t^*}(\tau, t).
\end{equation}
We give $\ell$ explicitly as \eqref{eq:site_pattern_profile_likelihood_mean} in the Appendix.
For a fixed $\tau$, let $\hat{t}_\nCols$ maximize the left-hand side of \eqref{eq:exp-log-lik} and $\hat{t}$ maximize the right-hand side.
We show in the Appendix that $\hat{t}_\nCols \rightarrow \hat{t}$, allowing us to focus on only the right-hand side above.

%EM Could we give an extra sentence justifying that the limit of the inferential process will be the same as the inferences applied to the limiting data? Compare https://paperpile.com/shared/HohiVH in which they are rather different.
%das: moved to appendix

@@

%EM Suggest "Our main theorem states that ... internal branch length zero ..."
Our main theorem states that given data generated on $\tau_1$ there exist true nonzero branch lengths such that the estimator $\hat{t}$ maximizing the right-hand side of \eqref{eq:exp-log-lik} has an internal branch of length zero.
\begin{restatable}{theorem}{topoInconsist}
Let $\tau^*=\tau_1$, $t^*=\{x^*, y^*, x^*, y^*, y^*\}$, and $t=\{x_1, y_1, x_2, y_2, w\}$ with $x_1, y_1, x_2, y_2, w > 0$.
There exists a set of $0 < x^*, y^* < 1$ such that the solution $\hat{t} := \{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\}$ given by
\[
\hat{t} = \arg\max_{t} \ \ell_{\tau^*,t^*}(\tau_1, t)
\]
has the property $\hat{w}\equiv 1$.
\end{restatable}
